---
title: "geo_facet with scatterplots: the relationship between local revenue and student reading proficiency"
description: |
  Scatterplots for each state illustrating the relationship between district local revenue (total and from property taxes) and student reading/language arts profiency.
author:
  - name: Amy Warnock
date: 2022-03-13
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(edld652)
library(tidyverse)
library(here)
library(janitor)
library(rio)
#install.packages("geofacet")
library(geofacet)
```

```{r load-data}
rlalea_10 <- get_data("EDFacts_rla_achievement_lea_2010_2019")

fiscal2010 <- get_data("NCES_CCD_fiscal_district_2010")

regions <- import(here("data", "us_census_bureau_regions_and_divisions.csv"), 
									setclass = "tbl_df") %>% 
	rename(`STNAME` = State,
				 STABBR = `State Code`) 
```

#### Amy
**Research Question:** 

What is the relationship between district funding (both total local revenue and local revenue from property taxes) and outcomes by state?

**Design choices:**

-  Scatterplots enable the audience to observe the strength and direction of the association between two variables, as well as the spread, number of observations, and potentially type of relationship (e.g., linear, quadratic).
-  Using `geo_facet()` organized the scatterplots in the approximate shape of the United States (instead of alphabetically). 
-  Improved cognitive load by minimizing graph elements, such as shading, axis breaks, and grid lines
-  Selected a muted color for geom_points that was cohesive with the colors in the other plots I created
-  Used transparency with `geom_point()` to allow the audience to see the density of observations
-  Clear, succinct, and polished title and labels
-  Log transformation of the x-axis so that the points weren't clustered at the lower end of the range
	
**Intended audience:** 

Researchers and policy makers. 

**Final Plots** 

```{r cleaning-wrangling}
# Selected columns of interest. Filtered so that rows with suppressed values (e.g., -9, -2) for key variables of interest weren't included. 

viz3_fiscal2010 <- fiscal2010 %>% 
  select(LEAID, NAME, STABBR, CENSUSID, V33,
         TOTALREV, TFEDREV, TSTREV, TLOCREV, T06) %>% 
  filter(V33 >= 0, TLOCREV >= 0, T06 >= 0) %>%
  rename_with(tolower) %>% 
  rename(totalstu = v33, tlocrevtaxes = t06) %>% 
  mutate(locrev_stu = tlocrev / totalstu,
         locrevtaxes_stu = tlocrevtaxes / totalstu)

# Remove Inf values

viz3_fiscal2010 <- viz3_fiscal2010 %>% 
  mutate_if(is.numeric, list(~na_if(., Inf)))

# Narrowed the RLA 2010 file down to variables of interest. Selected percent proficient variables across all grades (00) for race ethnicity, disability, English Language Learner status, and economically disadvantaged subgroups. Transformed variable names to be lowercase, transformed state names to be title case, and replaced the suppressed values (e.g., PS, n/a, etc.) with NA.

viz3_rlalea00 <- rlalea_10 %>% 
  select(YEAR, 
         STNAM, 
         FIPST,
         LEAID, 
         ALL_RLA00PCTPROF, 
         MAM_RLA00PCTPROF,
         MAS_RLA00PCTPROF,
         MBL_RLA00PCTPROF,
         MHI_RLA00PCTPROF,
         MTR_RLA00PCTPROF, 
         MWH_RLA00PCTPROF,
         CWD_RLA00PCTPROF,
         ECD_RLA00PCTPROF,
         LEP_RLA00PCTPROF) %>% 
  rename_with(tolower) %>% 
  mutate(stnam = str_to_title(stnam), 
         (across(all_rla00pctprof:lep_rla00pctprof, 
                 ~replace(., . %in% c("PS", 
                                      "n/a",	
                                      "LT50",	
                                      "LE5",	
                                      "LE20",	
                                      "LE10",	
                                      "GE99",	
                                      "GE95",	
                                      "GE90",	
                                      "GE80",	
                                      "GE50"), NA))))

# Next step was cleaning the percentage columns to change percentage ranges to average percentages. I used the method Daniel used on the course data webpage and applied it to all subgroups.

# All = across all students

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(all_rla00pctprof, c("all_lower", "all_upper"), sep = "-") %>% 
  mutate(
    all_upper = ifelse(is.na(all_upper), all_lower, all_upper),
    all_lower = as.numeric(all_lower),
    all_upper = as.numeric(all_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_all = mean(c(all_lower, all_upper))) %>% 
  ungroup()

# mam = American Indian/Alaska Native

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(mam_rla00pctprof, c("mam_lower", "mam_upper"), sep = "-") %>% 
  mutate(
    mam_upper = ifelse(is.na(mam_upper), mam_lower, mam_upper),
    mam_lower = as.numeric(mam_lower),
    mam_upper = as.numeric(mam_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_mam = mean(c(mam_lower, mam_upper))) %>% 
  ungroup()

# mas = Asian/Pacific Islander

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(mas_rla00pctprof, c("mas_lower", "mas_upper"), sep = "-") %>% 
  mutate(
    mas_upper = ifelse(is.na(mas_upper), mas_lower, mas_upper),
    mas_lower = as.numeric(mas_lower),
    mas_upper = as.numeric(mas_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_mas = mean(c(mas_lower, mas_upper))) %>% 
  ungroup()

# mbl = Black

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(mbl_rla00pctprof, c("mbl_lower", "mbl_upper"), sep = "-") %>% 
  mutate(
    mbl_upper = ifelse(is.na(mbl_upper), mbl_lower, mbl_upper),
    mbl_lower = as.numeric(mbl_lower),
    mbl_upper = as.numeric(mbl_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_mbl = mean(c(mbl_lower, mbl_upper))) %>% 
  ungroup()

# mhi = Hispanic/Latino

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(mhi_rla00pctprof, c("mhi_lower", "mhi_upper"), sep = "-") %>% 
  mutate(
    mhi_upper = ifelse(is.na(mhi_upper), mhi_lower, mhi_upper),
    mhi_lower = as.numeric(mhi_lower),
    mhi_upper = as.numeric(mhi_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_mhi = mean(c(mhi_lower, mhi_upper))) %>% 
  ungroup()

# mtr = Multiracial

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(mtr_rla00pctprof, c("mtr_lower", "mtr_upper"), sep = "-") %>% 
  mutate(
    mtr_upper = ifelse(is.na(mtr_upper), mtr_lower, mtr_upper),
    mtr_lower = as.numeric(mtr_lower),
    mtr_upper = as.numeric(mtr_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_mtr = mean(c(mtr_lower, mtr_upper))) %>% 
  ungroup()

# mwh = White

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(mwh_rla00pctprof, c("mwh_lower", "mwh_upper"), sep = "-") %>% 
  mutate(
    mwh_upper = ifelse(is.na(mwh_upper), mwh_lower, mwh_upper),
    mwh_lower = as.numeric(mwh_lower),
    mwh_upper = as.numeric(mwh_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_mwh = mean(c(mwh_lower, mwh_upper))) %>% 
  ungroup()

# cwd = children with disabilities

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(cwd_rla00pctprof, c("cwd_lower", "cwd_upper"), sep = "-") %>% 
  mutate(
    cwd_upper = ifelse(is.na(cwd_upper), cwd_lower, cwd_upper),
    cwd_lower = as.numeric(cwd_lower),
    cwd_upper = as.numeric(cwd_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_cwd = mean(c(cwd_lower, cwd_upper))) %>% 
  ungroup()

# ecd = economically disadvantaged

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(ecd_rla00pctprof, c("ecd_lower", "ecd_upper"), sep = "-") %>% 
  mutate(
    ecd_upper = ifelse(is.na(ecd_upper), ecd_lower, ecd_upper),
    ecd_lower = as.numeric(ecd_lower),
    ecd_upper = as.numeric(ecd_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_ecd = mean(c(ecd_lower, ecd_upper))) %>% 
  ungroup()

# lep = limited English proficiency (English Language Learner)

viz3_rlalea00 <- viz3_rlalea00 %>% 
  tidyr::separate(lep_rla00pctprof, c("lep_lower", "lep_upper"), sep = "-") %>% 
  mutate(
    lep_upper = ifelse(is.na(lep_upper), lep_lower, lep_upper),
    lep_lower = as.numeric(lep_lower),
    lep_upper = as.numeric(lep_upper)
    ) %>% 
  rowwise() %>% 
  mutate(meanpctprof_lep = mean(c(lep_lower, lep_upper))) %>% 
  ungroup()

# Got rid of the "_lower" and "_upper" percentage columns since they won't be needed

viz3_rlalea00 <- viz3_rlalea00 %>% 
  select(year, 
         stnam, 
         fipst,
         leaid,
         contains("meanpctprof"))

# Pivoted the dataset longer to have a column for subgroup and a column for mean percentage proficient. 

viz3_rlalea00_long <- viz3_rlalea00 %>%
  pivot_longer(
 		cols = contains("meanpctprof"),
 		names_to = "subgroup",
 		values_to = "meanpctprof",
 		names_prefix = "meanpctprof_") 

# Joined the long file with the cleaned/narrowed fiscal data file. Used an inner join because I'm only interested in LEAs that have both student proficiency and fiscal data. 

viz3_rla00long_fiscal_2010 <- inner_join(viz3_rlalea00_long, 
                                         viz3_fiscal2010, 
                                         by = "leaid")

# Merge with census data, which includes region

viz3_rla00long_fiscal_2010 <- left_join(viz3_rla00long_fiscal_2010,
                                               regions, 
                                               by = c("stnam" = "STNAME"))
```

```{r panelset}
xaringanExtra::use_panelset()
```

::: l-page

::::: {.panelset}

::: {.panel}

## Local Revenue x Reading Proficiency {.panel-name}
```{r panel-locrev, layout = "l-page", fig.height = 8, fig.width = 14}
viz3_rla00long_fiscal_2010 %>% 
  filter(subgroup == "all") %>% 
  filter(locrev_stu > 0) %>% 
  ggplot(aes(x = locrev_stu, y = meanpctprof)) +
  facet_geo(~stnam) +
  geom_point(color = "#446455", stroke = 0, alpha = .5) + 
  scale_x_log10(breaks = c(10, 1000, 100000),
                labels = c("$10", "$1k", "$100k")) +
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100),
                     labels = c("0%", "25%", "50%", "75%", "100%")) +
  labs(title = "Relationship Between Total Local Revenue and RLA Proficiency",
       y = "Average percent proficient",
       x = "Dollar per student",
       caption = "Note the scale of the x-axis. 
       Source: National Center for Education Statistics, 2010") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank()) 
```
:::

::: {.panel}

## Property Tax Revenue x Reading Proficiency {.panel-name}
```{r panel-locrevtaxes, layout = "l-page", fig.height = 8, fig.width = 14}
viz3_rla00long_fiscal_2010 %>% 
  filter(subgroup == "all") %>% 
  filter(locrevtaxes_stu > 0) %>% 
  ggplot(aes(x = locrevtaxes_stu, y = meanpctprof)) +
  facet_geo(~stnam) +
  geom_point(color = "#446455", stroke = 0, alpha = .5) + 
  scale_x_log10(breaks = c(10, 1000, 100000),
                labels = c("$10", "$1k", "$100k")) +
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100),
                     labels = c("0%", "25%", "50%", "75%", "100%")) +
  labs(title = "Relationship Between Local Revenue from Property Taxes and RLA Proficiency",
       y = "Average percent proficient",
       x = "Dollar per student",
       caption = "Note the scale of the x-axis. 
       Source: National Center for Education Statistics, 2010") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())
```

::: 

:::::

:::

**Findings and Observations:**

-  Overall, total local revenue and revenue from property taxes are positively associated with the reading/language arts outcomes of students. 
-  However, the strength of the association varies by state. 
-  There are states where there doesn't appear to be a relationship between the variables (e.g., Colorado, Texas, Vermont).
-  The relationship between local revenue and student reading proficiency appears to be strong in several states, including New Jersey. This state is notable as it has the highest average district local revenue from property taxes. Given more time and bandwidth, I would have liked to create a district-level choropleth of revenue in New Jersey (or even a bivariate map of property taxes and outcomes).

**Prior Versions** 

:::l-body-outset

::::: {.panelset}

::: {.panel}

## Prior Version 1 {.panel-name}

```{r panel-prior1, layout = "l-body-outset", fig.height = 10, fig.width = 10}
viz3_rla00long_fiscal_2010 %>% 
  filter(subgroup == "all") %>% 
  ggplot(aes(x = locrev_stu, y = meanpctprof)) +
  geom_point() 
```
:::

::: {.panel}
## Prior Version 2 {.panel-name}
```{r panel-prior2, layout = "l-body-outset", fig.height = 10, fig.width = 10}
viz3_rla00long_fiscal_2010 %>% 
  filter(subgroup == "all") %>% 
  ggplot(aes(x = locrev_stu, y = meanpctprof)) +
  geom_point(color = "gray30", fill = "gray30", alpha = .4) + 
  scale_x_log10(labels = scales::dollar) +
  labs(title = "Relationship between Local Revenue and RLA Proficiency",
       y = "Approximate Average Percent Proficient",
       x = "Dollar per Student",
       caption = "Source: National Center for Education Statistics, 2010") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())
```
:::

::: {.panel}
## Prior Version 3 {.panel-name}
```{r panel-prior3, layout = "l-body-outset", fig.height = 10, fig.width = 10}
viz3_rla00long_fiscal_2010 %>% 
  filter(subgroup == "all") %>% 
  ggplot(aes(x = locrev_stu, y = meanpctprof)) +
  geom_point(color = "#798E87", stroke = 0, alpha = .4) + 
  scale_x_log10(labels = scales::dollar) +
  labs(title = "Relationship between Local Revenue and RLA Proficiency",
       y = "Approximate Average Percent Proficient",
       x = "Dollar per Student",
       caption = "Source: National Center for Education Statistics, 2010") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())
```
:::

::: {.panel}
## Prior Version 4 {.panel-name}
```{r panel-prior4, layout = "l-body-outset", fig.height = 10, fig.width = 10}
viz3_rla00long_fiscal_2010 %>% 
  filter(subgroup == "all") %>% 
  filter(locrev_stu > 0) %>% 
  ggplot(aes(x = locrev_stu, y = meanpctprof)) +
  facet_wrap(~stnam) +
  geom_point(color = "gray30", fill = "gray30", alpha = .4) + 
  scale_x_log10(labels = scales::dollar) +
  labs(title = "Total Local Revenue and RLA Proficiency",
       y = "Approximate Average Percent Proficient",
       x = "Dollar per Student",
       caption = "Source: National Center for Education Statistics, 2010") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank()) 

```
:::

:::::

:::

**Evolution of plots over time:** 

As both plots are similar in regard to format and design choices and went through the same evolution, I've included the previous versions of just one plot above. 

-  I began with exploring the overall relationship across all states and districts. Prior Version 1 is my initial rough and basic attempt at this visualization Default settings and elements are used. The points are clustered near zero.
-  In Prior Version 2, I made some refinements (alpha, x-axis scale, minimal graph elements, etc.) 
-  In Prior Version 3, I decided to try color and added `stroke = 0` based on the helpful feedback from peer reviews. 
-  Given that policies of district funding and property taxes vary at the state level, I wanted to be sure to explore scatterplots at the state level. In this plot (Prior Version 4), I faceted by state. The upside of faceting by state is that only the states with data are included. However, given regional patterns observed with the bar plots and maps, I ended up with `facet_geo` so that the plots were roughly arranged in the shape of the US. I also added some color back in and adjusted breaks and labels. 
 



